Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems over Hilbert Spaces

### 2023-05-10 latex/intro.tex
Improving

\section{Introduction}\label{section:intro}
Measurements play a fundamental role in generating observations and
are indispensable to any inference process. In the realm of data
science, selecting the optimal set of measurements holds particular
significance when inferring parameters of a physical process. Unlike
many other fields, where measurements are fixed, in this context, we
have the freedom to choose which measurements to take. This freedom
should be taken advantage of to choose measurements that enhance
accuracy, reduce costs, or achieve both objectives
simultaneously. Whether measurements involve specifying electrode
locations on the skin in electric impedance tomography
\cite{horesh2010impedance}, determining certain wavelengths in MRI
\cite{horesh2008mri}, or positioning sensors for detecting
ground-reflected waves in the search for oil
\cite{horesh2008borehole}, the selection of optimal measurements,
referred to as the problem of \emph{optimal design}, becomes
crucial. To achieve this, researchers optimize specific \emph{design
criteria}, with A- and D-optimality being two of the most widely
recognized and extensively studied design criteria
\cite{Chaloner1995}.

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  hooker2009, fedorov2012, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in Section
\ref{section:example}). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. This intriguing phenomenon is
known as \emph{measurement clusterization} \cite{Ucinski05}, and we
refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

@@

In spatial problems involving choice of measuremnts within a domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$, many researchers circumvent
the problem of measurement clusterization by choosing measurements
from a coarse grid in $\Omega$ \cite{koval2020, alexanderian2021,
  attia2020, alexanderian2014, alexanderian2016,
  alexanderian2018efficient}. However, this approach incurs a
significant computational cost as it requires solving a difficult
combinatorial optimization problem for choosing measurementst from a
finite set. The combinatorial optimization problem is usually relaxed
by first assigning optimal measurement "weights" in $\mathbb{R}_+$ to
the potential measurement locations. Some researchers incorporate a
sparsifying $\ell_1$ penalty term into the design criterion, which is
subsequently thresholded to achieve the desired binary design over the
coarse grid \cite{horesh2008borehole}. Others progressively relax the
$\ell_1$ penalty to an $\ell_0$ penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. A clever alternative is to
cast the problem of finding optimal measurement weights as a
stochastic optimization problem \cite{attia2022stochastic}. All of the
aforementioned methods may indeed find a binary optimal design
restricted to a given coarse grid. However, none addresses one
fundamental issue: the restriction of measurement locations to a
coarse grid in $\Omega \subseteq \mathbb{R}^d, d=1,2,3$ inherently
results in a sub-optimal design.

Avoiding measurement clusterization is a pragmatic approach:
intuitively, we recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, we strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves this objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. To the best of my knowledge, no previous
study has tried to address some seemingly simple yet fundamental
questions:
%
Why does measurement clusterization occur?
%
Why does imposing correlations between observations alleviate
measurement clusterization?
%
Should we aim to avoid measurement clusterization?
%
Is it possible to substitute an optimal clustered design with an
equally optimal non-clustered design?
%Can an optimal clustered design be relpaced with an equally optimal
%non-clustered design?

\subsection{Contribution}
The primary objective of this study is to provide a comprehensive
understanding of measurement clusterization by addressing the
aforementioned questions. Our focus centers around investigating the
Bayesian D-optimality criterion, which involves maximizing the
expected Kullback-Leibler divergence between the posterior and prior
measures \cite{CoverThomas91, Chaloner1995}. We conduct an analysis of
Bayesian D-optimal designs within the context of linear inverse
problems in Hilbert spaces. We propose a novel relaxed model for
D-optimality that maintains analytical tractability and enables the
identification of D-optimal designs using Lagrange multipliers. This
analytical framework facilitates the exploration of the questions
posed at the end of the previous paragraph:

\begin{enumerate}
\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} In Section \ref{section:vanishing}, we provide an insightful
  explanation for the optimality of clustered designs when no model
  error is present. We demonstrate that measurement clusterization
  poses no obstruction to D-optimality and that clustered design are
  preferred by numerical algorithms.

@@

The cornerstone of our investigation into measurement clusterization
is Theorem \ref{thm:char} proven in Section
\ref{section:vanishing}. The key insight of Theorem \ref{thm:char}
lies in its fifth component, which highlights that a D-optimal design
aims to uniformly reduce posterior uncertainties across all posterior
covariance eigenvectors. A similar conclusion was reached by Koval et
al.~\cite{koval2020}, who showed that A-optimal designs should be
constructed in the space of \emph{observations}.

@@

Other contributions include three useful lemmas: In Lemma
\ref{lemma:free} we decompose a symmetric positive definite matrix $M
\in \mathbb{R}^{k \times k}$ with $\ttr M = m \in \mathbb{N}$ as $M =
AA^t$, where $A$ has unit norm columns. In Lemma \ref{lemma:MDL} we
prove a matrix determinant lemma in Hilbert spaces. Finally in Lemma
\ref{lemma:lax} we generalize a lemma for calculating
$\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
operator valued function \cite{Lax07}.

>>>>

\section{Introduction}\label{section:intro}
Measurements play a fundamental role in generating observations and
are indispensable to any inference process. In the realm of data
science, selecting the optimal set of measurements holds particular
significance when inferring parameters of a physical process. Unlike
many other fields, where measurements are fixed, in this context, we
have the freedom to choose which measurements to take. This freedom
should be taken advantage of: measurements should be chosen to enhance
accuracy, reduce costs, or both. Whether measurements involve
specifying electrode locations on the skin in electric impedance
tomography \cite{horesh2010impedance}, determining certain wavelengths
in MRI \cite{horesh2008mri}, or positioning sensors for detecting
ground-reflected waves in the search for oil
\cite{horesh2008borehole}, the selection of optimal measurements,
referred to as the problem of \emph{optimal design}, becomes
crucial. Optimal measurements are chosen according to specific
\emph{design criteria}, with A- and D-optimality being two of the most
widely recognized and extensively studied design criteria
\cite{Chaloner1995}.

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  hooker2009, fedorov2012, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in Section
\ref{section:example}). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. We refer to this intriguing
phenomenon as \emph{measurement clusterization} \cite{Ucinski05}, and
we refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

@@

In spatial problems involving choice of measuremnts within a domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$, many researchers circumvent
the problem of measurement clusterization by choosing measurements
from a coarse grid in $\Omega$ \cite{koval2020, alexanderian2021,
  attia2020, alexanderian2014, alexanderian2016,
  alexanderian2018efficient}. This approach incurs a significant
computational cost as it requires solving a difficult combinatorial
optimization problem for measurements locations over a finite set. The
combinatorial optimization problem is usually relaxed by first
assigning optimal measurement ``weights'' in $\mathbb{R}_+$ to the
potential measurement locations. Some researchers incorporate a
sparsifying $\ell_1$ penalty term into the design criterion, which is
subsequently thresholded to achieve the desired binary design over the
coarse grid \cite{horesh2008borehole}. Others progressively relax the
$\ell_1$ penalty to an $\ell_0$ penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. A clever alternative is to
cast the problem of finding optimal measurement weights as a
stochastic optimization problem \cite{attia2022stochastic}. All of the
aforementioned methods may indeed find a binary optimal design
restricted to a given coarse grid. However, none addresses one
fundamental issue: the restriction of measurement locations to a
coarse grid in $\Omega$ inherently results in a sub-optimal design.

Avoiding measurement clusterization is a pragmatic approach:
intuitively, we recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, we strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves its objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. To the best of my knowledge, no previous
study has tried to address some seemingly simple yet fundamental
questions:
%
Why does measurement clusterization occur?
%
Why does imposing correlations between observations alleviate
measurement clusterization?
%
Should we aim to avoid measurement clusterization?
%
Is it possible to substitute an optimal clustered design with an
equally optimal non-clustered design?
%Can an optimal clustered design be relpaced with an equally optimal
%non-clustered design?

\subsection{Contribution}
The primary objective of this study is to provide a comprehensive
understanding of measurement clusterization by addressing the
aforementioned questions. Our focus centers around investigating the
Bayesian D-optimality criterion, which involves maximizing the
expected Kullback-Leibler divergence between the posterior and prior
measures \cite{CoverThomas91, Chaloner1995}. We conduct an analysis of
Bayesian D-optimal designs within the context of linear inverse
problems over Hilbert spaces. We propose a novel relaxed model for
D-optimality that maintains analytical tractability and enables the
identification of D-optimal designs using Lagrange multipliers. This
analytical framework facilitates the exploration of the questions
posed at the end of the previous paragraph:

\begin{enumerate}
\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} In Section \ref{section:vanishing}, we provide an insightful
  explanation for the optimality of clustered designs when no model
  error is present. We demonstrate that measurement clusterization
  poses no obstruction to D-optimality and that clustered designs are
  preferred by numerical algorithms.

@@

The cornerstone of our investigation into measurement clusterization
is Theorem \ref{thm:char} proven in Section
\ref{section:vanishing}. The key insight of Theorem \ref{thm:char}
lies in its fifth component, which highlights that a D-optimal design
aims to uniformly reduce posterior uncertainties across all posterior
covariance eigenvectors. A similar conclusion was reached by Koval et
al.~\cite{koval2020}, who showed that A-optimal designs are best
constructed in the space of observations.

@@

Other contributions include three useful lemmas: In Lemma
\ref{lemma:free} we decompose a symmetric positive definite matrix $M
\in \mathbb{R}^{k \times k}$ with $\ttr M = m \in \mathbb{N}$ as $M =
AA^t$, where $A$ has unit norm columns. In Lemma \ref{lemma:MDL} we
prove a matrix determinant lemma in Hilbert spaces. Finally, in Lemma
\ref{lemma:lax} we generalize a lemma for calculating
$\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
operator valued function \cite{Lax07}.

### 2023-05-10 latex/paper.tex
Improving

\title{Measurement Clusterization in D-optimal Designs for Bayesian Linear
  Inverse Problems}

@@

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  We investigate the problem of measurement clusterization in D-optimal
  experimental design for infinite-dimensional Bayesian linear inverse
  problems. Many authors seem to believe that the fact that spatial
  correlations mitigateWe suggest an analytically tractable model for D-optimal
  designs. We argue that measurement clusterization in the case of iid
  observation noise is generic. We also prove that a model error term
  mitigates the clusterization phenomenon.

  %% As a part of the analysis we prove a Matrix Determinant
  %% Lemma in Hilbert spaces and a lemma for calculating
  %% $\frac{\der}{\der t} \log \det (I + X(t))$ for certain
  %% operator-valued functions. We also show how to decompose $M = AA^t$
  %% with $A$ constrained to have unit norm columns.
\end{abstract}

>>>>

\title{Measurement Clusterization in D-optimal Designs for Bayesian Linear
  Inverse Problems over Hilbert Spaces}

@@

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  Inversion for parameteres of physical processes typically requires
  taking expensive measurements. Finding an optimal set of
  measurements is known as the optimal design problem. Surprisingly,
  measurement locations in optimal design are sometimes extremely
  clustered. Practically and intuitively, measurement clusterization
  is a huge problem. Researchers usually avoid measurement
  clusterization by somehow modifying the optimal design problem. We
  consider a certain flavor of the optimal design problem, based on
  the Bayesian D-optimality criterion. We suggest an analytically
  tractable model for Bayesian D-optimal designs in linear inverse
  problems over Hilbert spaces. We prove that measurement
  clusterization poses no obstruction to D-optimality. We demonstrate
  that measurement clusterization is a generic property of D-optimal
  designs, and prove that measurement correlation mitigates the
  clusterization of measurements. We also give a full characterization
  of D-optimal designs, which shows that D-optimal designs uniformly
  reduce uncertainty in the space of observations. Most importantly,
  we show how measurement clusterization can be avoided without
  changing the optimal design problem.
\end{abstract}
